\cursedforest\ - A Random Forest Implementation for ``Big'' and ``Wide'' Data \\ Supplementary Information

### 2016-10-28 doc/publications/plos16/plos16.tex
small changes to plos16.tex and added  supplementary_information.tex

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB or RAM.  We captured the OOB (out of bag)
classification error and the 20 most important variables.

Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l+l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & RF - Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\

@@

                                                    & RF - Spark ML  & 0.94       & 8hr 29min        & 8GB \\
                                                    & LR - Spark ML  &            & 2hr 6min         & 8GB \\

@@

                                                    & LR - Spark ML  &            & 14hr 48min                                           & 8GB \\

@@

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
& RF - Spark ML & - & - & - \\
& LR - Spark ML & - & - & - \\ 
& kmeans - Spark ML  & - (0.82)      & 30hr 44min    & 24GB \\ 
& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA     &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\
\hline
% \begin{tabular}{|p{1.2cm}|p{1cm}|p{1.5cm}+l|l|p{1.8cm}|p{1.5cm}|}
% \hline
% %\multicolumn{4}{|l|}{\bf Heading1} & \multicolumn{4}{|l|}{\bf Heading2}\\ \hline
% \bf{Data- set} & \bf{Samp- les} & \bf{Feat- ures}  & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/ Node} \\
% \hline
% %\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & CursedForest & 0.077 (XX)  &  &  \\
% phase1 chr22 &1,092 & 490,036 & CursedForest & 0.077 (XX)  & 9min 16sec & 1GB\footnote{\label{note10}10 nodes} \\
% &&& RF - Spark ML & 0.069 (0.89) & 14min 13sec & 1GB\footref{note10}  \\
% &&& RF - R &  0.05  & 24min  & 18GB\footref{note1}\footnote{using the \texttt{parallel} packages}\\
% &&& ranger -R &         0.05 &        1min 10sec  &          16GB  (one processor) \\
%  \hline
% phase3 chr22 & 2,504 & 1,103,548 &  &  &  &  \\
% &&& CursedForest & 0.053 & 9 min 8 sec & 2GB\footnote{\label{note20}20 nodes}\\
% &&& RF - Spark ML & 0.058 (0.92) & 1hr 1min & 2GB\footref{note20} \\
% &&& RF - R & NA  & NA & NA\footnote{long vectors ($> 2^31-1$)  are not supported in the .Fortran interface}\\
% &&&  H2O   &        0.05 &     16 hours       &   600GB \footnote{java -Xmx600g -jar h2o.jar,   h2o.init(nthreads=20)} \\
% \hline
% phase3 chr1 & 2,504 & 6,450,364 & CursedForest & 0.063 (XX) & 1hr 58min & 4GB \\
% &&& RF - Spark ML & (0.94) & 8hr 29min & 8GB \\
% &&& LR - Spark ML&  & 2hr 6min & 8GB\\ 
% &&& ranger  R   &  0.06   &     28 mins  &            num.threads = 10\footnote{don't know how to get the memory used, This
%   was done on dpas-03} \\ 
% \hline
% phase3 chr1-3 & 2,504 & 19,328,051  & CursedForest & 0.046 (XX) & 4hr 0min & 4GB\\
% &&& RF - Spark ML & - & -  & -\\
% &&& LR - Spark ML & & 14hr 48min & 8GB \\ 
% &&& ranger  R   &  0.05   &     58 mins\footnote{ but took 6 hours to load the data}  &      909GB\footnote{I think this is total memory. The job was run with
%   ntasks-per-node=160 cores-per-socket=10 on ruby }  \\ 
% \hline
% phase3 chr1-22 & 2,504 & 81,047,467 & CursedForest & XX (0.96) & 7hr 14min & 8GB \\ 
% &&& RF - Spark ML & - & - & - \\
% &&& LR - Spark ML & - & - & - \\ 
% %&&& kmeans - Spark ML & - (0.82) & 30hr 44min & 24GB \\ 
% \hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% Section 1:  Feature Selection and classification on synthetic data
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{\cursedforest\ can evaluate variable importance for millions of features}\label{synthetic}
%In this section we demonstrate that \cursedforest's ability to rapidly grow large number of trees, \ntree, in
%parallel results in a stable feature importance ranking.  This concept was originally proposed by Liaw and
%Weiner~\cite{Liaw.and.Weiner.2002}, who suggested that though the variable importance measures may vary from run to
%run, the ranking of the variables will stabilise with growing \ntree.
In this section we demonstrate \cursedforest\ two main features, namely variable importance analysis to select the most
associated set of features amongst millions of dimensions as well as its ability to accurately classify the labels of
high-dimensional samples.
%We therefore generate a synthetic dataset with varying numbers of variables (features) and samples, as discussed in the method section. 
We therefore generate a synthetic dataset with 2.5 million variables (features) and 5000 samples, as discussed in the
method section.  Briefly we generate features by randomly sampling from a three value vector (1,2,3) with equal
\marginpar{repicated methods section}
probabilities, which mimics genomic mutation data and corresponds to homozygote reference, heterozygote or homozygote
alternative.  We then generate the response variable as a linear combination of 5 weighted features, modulating a
complex trait of multiple interacting features.

After fitting the random forest model and performing the variable importance analysis we hence expect to retrieve these
5 feature in order of the weight that was assigned to them.  We are running Apache Spark 1.6.1 on a YARN cluster with 12
worker nodes each with 16 Intel Xeon E5-2660@2.20GHz CPU cores and 128 GB of RAM.  We captured the OOB (out of bag)
classification error and the 20 most important variables.

\marginpar{the accuracy hardly changes}
Figure~\ref{figure:synth}a illustrates that increasing the number of predictors sampled for splitting at each node
(\mtry) increases the accuracy with which features are ranked.  From $mtry=0.5$ the sort order of the features reflects
the weight originally assigned to them (dotted line). Further increasing \mtry\ improves the predicted importance
measure, which starts to converge to the weight originally assigned to them (coefficients).  It also increases the
difference between the associated 5 variable and the remaining variables in the top 20 list.  By $mtry=0.8$ the variable
with the smallest weight has a predicted importance measure that is 5 times larger than the aggregated importance of the
15 remaining variables from the list.

@@

\begin{table}[!ht]
\begin{minipage}{\textwidth}
\begin{adjustwidth}{-2.25in}{0in} % Comment out/remove adjustwidth environment if table fits in text column.
\centering
\caption{
{\bf Performance comparison between the different machine learning algorithms.}}
\begin{tabular}{|l|l|l|l|l|l|p{1cm}|}
\hline
\bf{Genome \%}                      & \bf{Method} & \bf{Error (ARI)} & \bf{Runtime} & \bf{Memory/Node} \\
\hline

\multirow{3}{*}{phase1\_chr22: 1,092 x 490,036} & \cursedforest\ & 0.077 (XX) &                  &                                                                   \\
                                                & Spark ML  &            &                  &                                                                   \\
                                                & RF - R         & 0.05       & 24 min           & 18GB\footnote{using the \texttt{parallel} packages with 10 nodes} \\
                                                & ranger - R      & 0.94       & 1 min 10 seconds & 16GB  (one processor)                                             \\
                                                & ranger - c++     & 0.94       & 7 mins           & --nthreads 20                                                     \\

@@

                                                    & Spark ML  & 0.94       & 8hr 29min        & 8GB \\

@@

@@

\multirow{4}{*}{phase3\_chr1-22: 2,504 x 81,047,467} & \cursedforest\ & XX (0.96) & 7hr 14min & 8GB \\
& Spark ML & - & - & - \\
& ranger (C++)       &        NA     &        NA     &            80 CPU -- runs out of memory (12.8GB per CPU) \\
& ranger (C++)       &        NA     &        NA     &            160
                                                        CPU -- runs
                                                        out of time at
                                                        20 hours \\
\hline
\end{tabular}
\begin{flushleft} 
Comparison of different various machine learning libraries on different subsets of variant data 
from the 1000 Genomes Project.
Each random forest model consists of 50 trees.
\end{flushleft}
\label{table1}
\end{adjustwidth}
\end{minipage}
\end{table}
